---
title: "SDS 315 HW 2"
author: "Ryan Paul Clarance"
date: "2025-01-28"
output: pdf_document
---

GitHub Link : https://github.com/RPC777/Homework-2

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(ggplot2)
library(dplyr)
library(readr)
library(tidyverse)
bikeshare <- read_csv("Desktop/SDS 313/bikeshare.csv")
head(bikeshare)
profs <- read_csv("Desktop/SDS 313/profs.csv")
capmetro_UT <- read_csv("Desktop/SDS 313/capmetro_UT.csv")
head(capmetro_UT)
head(profs)
billboard <- read_csv("Desktop/SDS 313/billboard.csv")
head(billboard)
```

<span style="font-family: Arial; font-size: 14pt; color: red;">
Question 1:


<span style="font-family: Arial; font-size: 14pt; color: red;">
Part A: Histogram

```{r}

ggplot(profs, aes(x = eval)) +
  geom_histogram(binwidth = 0.25, fill = "steelblue", color = "black") +
  labs(
    title = "Course Evaluation Scores Distribution",
    x = "Evaluation Score",
    y = "Frequency"
  ) +
  theme_minimal()
```
*Figure 1: This histogram shows how the values of the numeric variable are spread out. The taller bars indicate the most common values in the dataset. If the bars are mostly concentrated on one side, the data may be skewed. If the distribution is symmetric, it suggests a normal distribution. Understanding this spread helps in identifying trends and outliers.*


\newpage

<span style="font-family: Arial; font-size: 14pt; color: red;">
Part B: Boxplots 

```{r}
ggplot(profs, aes(x = native, y = eval, fill = native)) +
  geom_boxplot() +
  labs(
    title = "Course Evaluation Scores by Native English Speaker Status",
    x = "Native English Speaker",
    y = "Evaluation Score"
  ) +
  scale_fill_manual(values = c("yes" = "coral", "no" = "cyan")) +
  theme_minimal()
```
*Figure 2: This boxplot helps understand the range and spread of the numeric data. The box shows the middle 50% of the data, with the median marked by a line inside. The whiskers show the overall range, while dots outside them indicate values that are significantly higher or lower than the rest. This visualization helps in spotting anomalies and understanding vairables individually.*


\newpage


<span style="font-family: Arial; font-size: 14pt; color: red;">
Part C: Histograms

```{r}
ggplot(profs, aes(x = eval, fill = gender)) +
  geom_histogram(binwidth = 0.25, color = "black", alpha = 0.7) +
  facet_wrap(~ gender, nrow = 2) +
  labs(
    title = "Distribution of Course Evaluation Scores based on Gender",
    x = "Evaluation Score",
    y = "Frequency"
  ) +
  scale_fill_manual(values = c("male" = "blue", "female" = "pink")) +
  theme_minimal()
```
*Figure 3: This histogram shows the distribution of the evaluation scores for male and female professors. The x-axis shows the scores, while the y-axis shows the frequency of each score. The two separate facets allow us to compare the distributions for each gender. If the bars are more concentrated towards the higher or lower end, it suggests a trend in how students rate professors based on gender. This visualization helps in identifying any noticeable patterns or disparities in evaluations.*


\newpage


<span style="font-family: Arial; font-size: 14pt; color: red;">
Part D: Scatterplots

```{r}
ggplot(profs, aes(x = beauty, y = eval)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Relationship Between Physical Attractiveness and Evaluation Scores",
    x = "Physical Attractiveness (Standardized)",
    y = "Evaluation Score"
  ) +
  theme_minimal()
```
*Figure 4: This scatter plot shows how the physical attractiveness of a professor relates to their evaluation scores. Each point represents a professor, with their attractiveness score on the x-axis and their evaluation score on the y-axis. The red line shows the trend—if it slopes upward, it means higher attractiveness is linked to higher evaluations. The spread of points helps us see whether this relationship is strong or weak.*


\newpage

<span style="font-family: Arial; font-size: 14pt; color: red;">
Question 2:

<span style="font-family: Arial; font-size: 14pt; color: red;">
Plot A:

```{r}
plot_a_data <- bikeshare |>
  group_by(hr) |>
  summarize(avg_rentals = mean(total, na.rm = TRUE))

# Create Plot A
ggplot(plot_a_data, aes(x = hr, y = avg_rentals)) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  labs(
    title = "Plot A: Average Hourly Bike Rentals",
    x = "Hour of the Day",
    y = "Average Rentals"
  ) +
  theme_minimal()
```

*This scatterplot shows the average number of bike rentals by hour of the day for all days in the dataset. The x-axis represents the hour of the day from 0-23, and the y-axis represents the average number of rentals. A clear trend is shown, with peak rentals during commuting hours that are from 8–9 AM and 5–6 PM.*

\newpage



```{r}
plot_b_data <- bikeshare |>
  group_by(hr, workingday) |>
  summarize(avg_rentals = mean(total, na.rm = TRUE))

# Create Plot B
ggplot(plot_b_data, aes(x = hr, y = avg_rentals, color = factor(workingday))) +
  geom_line() +
  facet_wrap(~ workingday, labeller = labeller(workingday = c("0" = "Non-Working Day", "1" = "Working Day"))) +
  labs(
    title = "Plot B: Average Hourly Bike Rentals based on Working Day",
    x = "Hour",
    y = "Average Rentals",
    color = "Working Day"
  ) +
  theme_minimal()
```

*This  plot shows the average hourly bike rentals based on working and non-working days. Each panel represents a category of working day (working or non-working). The x-axis represents the hour of the day from 0-23, and the y-axis is the average number of rentals. On working days, rental peaks occur during commuting hours from 8–9 AM and 5–6 P), while on non-working days, based on the graph, rentals are more evenly distributed, with a rise around mid-morning*

\newpage

<span style="font-family: Arial; font-size: 14pt; color: red;">
Plot C:

```{r}
# Filter data for 9 AM and calculate averages by weather and working day
plot_c_data <- bikeshare |>
  filter(hr == 9) |>
  group_by(weathersit, workingday) |>
  summarize(avg_rentals = mean(total, na.rm = TRUE))

ggplot(plot_c_data, aes(x = factor(weathersit), y = avg_rentals, fill = factor(weathersit))) +
  geom_bar(stat = "identity") +
  facet_wrap(~ workingday, labeller = labeller(workingday = c("0" = "Non-Working Day", "1" = "Working Day"))) +
  labs(
    title = "Plot C: Rentals at 9 AM by Weather and Working Day",
    x = "Weather Condition",
    y = "Average Rentals",
    fill = "Weather"
  ) +
  scale_x_discrete(labels = c("1" = "Clear", "2" = "Mist", "3" = "Light Snow/Rain", "4" = "Heavy Rain/Snow")) +
  theme_minimal()
```

*This faceted bar plot shows the average number of rentals at 9 AM, grouped by weather conditions on the x-axis and categorized by working or non-working days. Weather conditions range from clear (1) to heavy rain/snow (4). When comparing working days to non-working days, Rentals are higher in better weather.*

\newpage


<span style="font-family: Arial; font-size: 14pt; color: red;">
Question 3:

```{r}
capmetro_UT <- capmetro_UT |>
  mutate(
    day_of_week = factor(day_of_week, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
    month = factor(month, levels = c("Sep", "Oct", "Nov"))
  )
```



Part 1: Line Graph for Average Boardings

```{r}
# Wrangle data for average boardings
avg_boardings <- capmetro_UT |>
  group_by(day_of_week, month, hour_of_day) |>
  summarize(avg_boardings = mean(boarding, na.rm = TRUE))

# Plot the faceted line graph
ggplot(avg_boardings, aes(x = hour_of_day, y = avg_boardings, color = month)) +
  geom_line(size = 1) +
  facet_wrap(~day_of_week, scales = "free_y") +
  labs(
    title = "Average Boardings",
    x = "Hour of Day",
    y = "Average Boardings",
    color = "Month"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 10)
  )
```


*This line graph shows the average number of boardings at different hours of the day, broken down by day of the week and month. The busiest times are usually in evening from 5-6pm and sometimes in the morning, likely due to students commuting to and from campus. Mondays in September have fewer boardings, possibly because fewer students are on campus at the start of the semester. There’s also a drop in boardings from Wednesday to Friday in November, which could be due to students leaving early for Thanksgiving break or a reduced bus schedule*

\newpage


Part 2:

```{r}
# Add weekend status for scatter plot
capmetro_UT <- capmetro_UT |>
  mutate(weekend_status = ifelse(weekend == 1, "Weekend", "Weekday"))

# Plot the faceted scatter plot
ggplot(capmetro_UT, aes(x = temperature, y = boarding, color = weekend_status)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~hour_of_day, scales = "free_y") +
  labs(
    title = "Boardings vs. Temperature by Hour of Day and Weekend Status",
    x = "Temperature (°F)",
    y = "Boardings",
    color = "Weekend Status"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 10)
  )
```

*Temperature Effect on Ridership is where Temperature has an effect on ridership, especially in the early morning and late evening hours, with higher temperatures correlating with increased boardings.The effect of temperature is more on weekdays, as weekday riding  follows commuting patterns.*


\newpage



```{r}
billboard_data <- billboard |>
  mutate(
    year = as.numeric(year),
    week_position = as.numeric(week_position)
  )
```



<span style="font-family: Arial; font-size: 14pt; color: red;">
Question 4:

<span style="font-family: Arial; font-size: 14pt; color: red;">
Part A: Top 10 Most Popular Songs

```{r}
# Calculate total weeks for each song
top_songs <- billboard_data |>
  group_by(performer, song) |>
  summarize(total_weeks = n(), .groups = "drop") |>
  arrange(desc(total_weeks)) |>
  slice(1:10)

# Display the table
knitr::kable(
  top_songs,
  col.names = c("Performer", "Song", "Total Weeks"),
  
)
```

*Table 1: This table lists the top 10 most popular songs based on their total weeks on the Billboard Top 100 since 1958. This table contains the performer, name of the song, and total amount of weeks it was on the chart.*

\newpage


<span style="font-family: Arial; font-size: 14pt; color: red;">
Part B: Diversity in Music 

```{r}
# Filter data and count unique songs per year
musical_diversity <- billboard_data |>
  filter(year > 1958 & year < 2021) |>
  group_by(year) |>
  summarize(unique_songs = n_distinct(song), .groups = "drop")

# Plot musical diversity over time
ggplot(musical_diversity, aes(x = year, y = unique_songs)) +
  geom_line(color = "steelblue", size = 1) +
  labs(
    title = "Musical Diversity Over Time",
    x = "Year",
    y = "Number of Unique Songs",
  ) +
  theme_minimal()
```

*Figure 1: This line graph shows the number of unique songs that appeared on the Billboard Top 100 each year from 1959 to 2020. It shows a general increase in musical diversity over time, meaning more different songs were making it onto the charts. This trend suggests that the music industry has become more varied, with a wider range of artists and styles gaining popularity throughout the years.*

\newpage


<span style="font-family: Arial; font-size: 14pt; color: red;">
Part C: Artists with the Most Ten-Week Hits

```{r}
# Identify ten-week hits
ten_week_hits <- billboard_data |>
  group_by(performer, song) |>
  summarize(total_weeks = n(), .groups = "drop") |>
  filter(total_weeks >= 10)

# Count ten-week hits per artist
artists_hits <- ten_week_hits |>
  group_by(performer) |>
  summarize(ten_week_hits_count = n(), .groups = "drop") |>
  filter(ten_week_hits_count >= 30) |>
  arrange(desc(ten_week_hits_count))

# Plot bar chart for artists with the most ten-week hits
ggplot(artists_hits, aes(x = fct_reorder(performer, ten_week_hits_count), y = ten_week_hits_count)) +
  geom_bar(stat = "identity", fill = "darkorange") +
  coord_flip() +
  labs(
    title = "Artists with the Most Ten-Week Hits",
    x = "Artist",
    y = "Number of Ten-Week Hits",
    
  ) +
  theme_minimal()
```


*Figure 2:  This bar chart shows the 19 artists with at least 30 songs that stayed on the Billboard Top 100 for ten or more weeks since 1958. It highlights their lasting success, strong fan support, and influence in the music industry. These artists have consistently released songs that remained popular for a long time, showing their ability to connect with audiences over the years.*



