---
title: "SDS 315 HW 2"
author: "Ryan Paul Clarance"
date: "2025-01-28"
output: pdf_document
---

GitHub Link : https://github.com/RPC777/Homework-2

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(ggplot2)
library(dplyr)
library(readr)
library(tidyverse)
bikeshare <- read_csv("Desktop/SDS 313/bikeshare.csv")
head(bikeshare)
profs <- read_csv("Desktop/SDS 313/profs.csv")
capmetro_UT <- read_csv("Desktop/SDS 313/capmetro_UT.csv")
head(capmetro_UT)
head(profs)
billboard <- read_csv("Desktop/SDS 313/billboard.csv")
head(billboard)
```

<span style="font-family: Arial; font-size: 14pt; color: red;">
Problem 1:


<span style="font-family: Arial; font-size: 14pt; color: red;">
Part A: Histogram of Course Evaluation Scores

```{r}

ggplot(profs, aes(x = eval)) +
  geom_histogram(binwidth = 0.25, fill = "steelblue", color = "black") +
  labs(
    title = "Distribution of Course Evaluation Scores",
    x = "Evaluation Score",
    y = "Frequency"
  ) +
  theme_minimal()
```
*Figure 1: This histogram shows how the values of the numeric variable are spread out. The taller bars indicate the most common values in the dataset. If the bars are mostly concentrated on one side, the data may be skewed. If the distribution is symmetric, it suggests a normal distribution. Understanding this spread helps in identifying trends and outliers.*


\newpage

<span style="font-family: Arial; font-size: 14pt; color: red;">
Part B: Boxplots by Native English Speaker Status

```{r}
ggplot(profs, aes(x = native, y = eval, fill = native)) +
  geom_boxplot() +
  labs(
    title = "Course Evaluation Scores by Native English Speaker Status",
    x = "Native English Speaker",
    y = "Evaluation Score"
  ) +
  scale_fill_manual(values = c("yes" = "coral", "no" = "cyan")) +
  theme_minimal()
```
*Figure 2: This boxplot helps understand the range and spread of the numeric data. The box shows the middle 50% of the data, with the median marked by a line inside. The whiskers show the overall range, while dots outside them indicate values that are significantly higher or lower than the rest. This visualization helps in spotting anomalies and understanding variability.*


\newpage


<span style="font-family: Arial; font-size: 14pt; color: red;">
Part C: Faceted Histograms by Gender

```{r}
ggplot(profs, aes(x = eval, fill = gender)) +
  geom_histogram(binwidth = 0.25, color = "black", alpha = 0.7) +
  facet_wrap(~ gender, nrow = 2) +
  labs(
    title = "Distribution of Course Evaluation Scores by Gender",
    x = "Evaluation Score",
    y = "Frequency"
  ) +
  scale_fill_manual(values = c("male" = "blue", "female" = "pink")) +
  theme_minimal()
```
*Figure 3: This histogram displays the distribution of course evaluation scores for male and female professors. The x-axis represents the evaluation scores, while the y-axis shows how frequently each score appears. The two separate facets allow us to compare the distributions for each gender. If the bars are more concentrated towards the higher or lower end, it suggests a trend in how students rate professors based on gender. This visualization helps in identifying any noticeable patterns or disparities in evaluations.*


\newpage


<span style="font-family: Arial; font-size: 14pt; color: red;">
Part D: Scatterplot of Physical Attractiveness and Evaluation Scores

```{r}
ggplot(profs, aes(x = beauty, y = eval)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Relationship Between Physical Attractiveness and Evaluation Scores",
    x = "Physical Attractiveness (Standardized)",
    y = "Evaluation Score"
  ) +
  theme_minimal()
```
*Figure 4: This scatter plot shows how a professor’s physical attractiveness relates to their evaluation scores. Each point represents a professor, with their attractiveness score on the x-axis and their evaluation score on the y-axis. The red line shows the trend—if it slopes upward, it means higher attractiveness is linked to higher evaluations. The spread of points helps us see whether this relationship is strong or weak.*


\newpage

<span style="font-family: Arial; font-size: 14pt; color: red;">
Problem 2:

<span style="font-family: Arial; font-size: 14pt; color: red;">
Plot A:

```{r}
plot_a_data <- bikeshare |>
  group_by(hr) |>
  summarize(avg_rentals = mean(total, na.rm = TRUE))

# Create Plot A
ggplot(plot_a_data, aes(x = hr, y = avg_rentals)) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  labs(
    title = "Plot A: Average Hourly Bike Rentals (Across All Days)",
    x = "Hour of the Day",
    y = "Average Rentals"
  ) +
  theme_minimal()
```

*This plot shows the average number of bike rentals by hour of the day across all days in the dataset. The x-axis represents the hour of the day (0–23), and the y-axis represents the average number of rentals. A clear trend is shown, with peak rentals during commuting hours (8–9 AM and 5–6 PM).*

\newpage


<span style="font-family: Arial; font-size: 14pt; color: red;">
Plot B:
```{r}
plot_b_data <- bikeshare |>
  group_by(hr, workingday) |>
  summarize(avg_rentals = mean(total, na.rm = TRUE))

# Create Plot B
ggplot(plot_b_data, aes(x = hr, y = avg_rentals, color = factor(workingday))) +
  geom_line() +
  facet_wrap(~ workingday, labeller = labeller(workingday = c("0" = "Non-Working Day", "1" = "Working Day"))) +
  labs(
    title = "Plot B: Average Hourly Bike Rentals by Working Day",
    x = "Hour of the Day",
    y = "Average Rentals",
    color = "Working Day"
  ) +
  theme_minimal()
```

*This  plot shows the average hourly bike rentals for working and non-working days. Each panel represents a category of working day (working or non-working). The x-axis is the hour of the day (0–23), and the y-axis is the average number of rentals. On working days, rental peaks occur during typical commuting hours (8–9 AM and 5–6 PM), while on non-working days, rentals are more evenly distributed, with a peak around mid-morning*

\newpage

<span style="font-family: Arial; font-size: 14pt; color: red;">
Plot C:

```{r}
# Filter data for 9 AM and calculate averages by weather and working day
plot_c_data <- bikeshare |>
  filter(hr == 9) |>
  group_by(weathersit, workingday) |>
  summarize(avg_rentals = mean(total, na.rm = TRUE))

ggplot(plot_c_data, aes(x = factor(weathersit), y = avg_rentals, fill = factor(weathersit))) +
  geom_bar(stat = "identity") +
  facet_wrap(~ workingday, labeller = labeller(workingday = c("0" = "Non-Working Day", "1" = "Working Day"))) +
  labs(
    title = "Plot C: Rentals at 9 AM by Weather and Working Day",
    x = "Weather Situation",
    y = "Average Rentals",
    fill = "Weather"
  ) +
  scale_x_discrete(labels = c("1" = "Clear", "2" = "Mist", "3" = "Light Snow/Rain", "4" = "Heavy Rain/Snow")) +
  theme_minimal()
```

*This faceted bar plot shows the average number of rentals at 9 AM, grouped by weather conditions (x-axis) and categorized by working or non-working days. Weather conditions range from clear (1) to heavy rain/snow (4). Rentals are higher in better weather and on working days compared to non-working days.*

\newpage


<span style="font-family: Arial; font-size: 14pt; color: red;">
Problem 3:

```{r}
capmetro_UT <- capmetro_UT |>
  mutate(
    day_of_week = factor(day_of_week, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
    month = factor(month, levels = c("Sep", "Oct", "Nov"))
  )
```



Question 1: Faceted Line Graph for Average Boardings

```{r}
# Wrangle data for average boardings
avg_boardings <- capmetro_UT |>
  group_by(day_of_week, month, hour_of_day) |>
  summarize(avg_boardings = mean(boarding, na.rm = TRUE))

# Plot the faceted line graph
ggplot(avg_boardings, aes(x = hour_of_day, y = avg_boardings, color = month)) +
  geom_line(size = 1) +
  facet_wrap(~day_of_week, scales = "free_y") +
  labs(
    title = "Average Boardings by Hour of Day, Day of Week, and Month",
    x = "Hour of Day",
    y = "Average Boardings",
    color = "Month"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 10)
  )
```


*This line graph shows the average number of boardings at different hours of the day, broken down by day of the week and month. The busiest times are usually in the morning (8–9 AM) and evening (5–6 PM), likely due to students commuting to and from campus. Mondays in September have noticeably fewer boardings, possibly because fewer students are on campus at the start of the semester. There’s also a drop in boardings from Wednesday to Friday in November, which could be due to students leaving early for Thanksgiving break or a reduced bus schedule*

\newpage


Question 2:

```{r}
# Add weekend status for scatter plot
capmetro_UT <- capmetro_UT |>
  mutate(weekend_status = ifelse(weekend == 1, "Weekend", "Weekday"))

# Plot the faceted scatter plot
ggplot(capmetro_UT, aes(x = temperature, y = boarding, color = weekend_status)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~hour_of_day, scales = "free_y") +
  labs(
    title = "Boardings vs. Temperature by Hour of Day and Weekend Status",
    x = "Temperature (°F)",
    y = "Boardings",
    color = "Weekend Status"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 10)
  )
```

*Temperature Effect on Ridership is where Temperature has an effect on ridership, especially in the early morning and late evening hours, with higher temperatures correlating with increased boardings.The effect of temperature is more on weekdays, as weekday riding  follows commuting patterns.*


\newpage



```{r}
billboard_data <- billboard |>
  mutate(
    year = as.numeric(year),
    week_position = as.numeric(week_position)
  )
```



<span style="font-family: Arial; font-size: 14pt; color: red;">
Problem 4:

<span style="font-family: Arial; font-size: 14pt; color: red;">
Part A: Top 10 Most Popular Songs by Total Weeks on Billboard Top 100

```{r}
# Calculate total weeks for each song
top_songs <- billboard_data |>
  group_by(performer, song) |>
  summarize(total_weeks = n(), .groups = "drop") |>
  arrange(desc(total_weeks)) |>
  slice(1:10)

# Display the table
knitr::kable(
  top_songs,
  col.names = c("Performer", "Song", "Total Weeks"),
  caption = "Top 10 most popular songs based on total weeks on Billboard Top 100 since 1958."
)
```

*Table 1: This table lists the top 10 most popular songs based on their total weeks on the Billboard Top 100 since 1958. It includes the performer, song title, and total weeks on the chart.*

\newpage


<span style="font-family: Arial; font-size: 14pt; color: red;">
Part B: Musical Diversity Over Time

```{r}
# Filter data and count unique songs per year
musical_diversity <- billboard_data |>
  filter(year > 1958 & year < 2021) |>
  group_by(year) |>
  summarize(unique_songs = n_distinct(song), .groups = "drop")

# Plot musical diversity over time
ggplot(musical_diversity, aes(x = year, y = unique_songs)) +
  geom_line(color = "steelblue", size = 1) +
  labs(
    title = "Musical Diversity of Billboard Top 100 Over Time",
    x = "Year",
    y = "Number of Unique Songs",
    caption = "This figure shows the number of unique songs on the Billboard Top 100 per year (1959–2020)."
  ) +
  theme_minimal()
```

*Figure 1: This line graph illustrates the number of unique songs appearing on the Billboard Top 100 for each year from 1959 to 2020. A trend of increasing musical diversity is observed.*

\newpage


<span style="font-family: Arial; font-size: 14pt; color: red;">
Part C: Artists with the Most Ten-Week Hits

```{r}
# Identify ten-week hits
ten_week_hits <- billboard_data |>
  group_by(performer, song) |>
  summarize(total_weeks = n(), .groups = "drop") |>
  filter(total_weeks >= 10)

# Count ten-week hits per artist
artists_hits <- ten_week_hits |>
  group_by(performer) |>
  summarize(ten_week_hits_count = n(), .groups = "drop") |>
  filter(ten_week_hits_count >= 30) |>
  arrange(desc(ten_week_hits_count))

# Plot bar chart for artists with the most ten-week hits
ggplot(artists_hits, aes(x = fct_reorder(performer, ten_week_hits_count), y = ten_week_hits_count)) +
  geom_bar(stat = "identity", fill = "darkorange") +
  coord_flip() +
  labs(
    title = "Artists with the Most Ten-Week Hits",
    x = "Artist",
    y = "Number of Ten-Week Hits",
    caption = "This figure shows the 19 artists with at least 30 ten-week hits on the Billboard Top 100 since 1958."
  ) +
  theme_minimal()
```


*Figure 2: This bar chart displays the 19 artists with at least 30 ten-week hits on the Billboard Top 100 since 1958, showcasing their significant impact on the music industry.*



